---
title: "If using the code for: Growth categories and birth weight, within-animal growth response, and projection to market weight, please cite the paper"
subtitle: >-
  Manuscript: *Increasing the threonine-to-lysine ratio improves the growth
  performance of slow-growing pigs within group precision feeding*. Submitted
  to the *Journal of Animal Science* (2026).
author: "Dr Aline Remus"
date: "`r format(Sys.Date(), '%d %B %Y')`"
version: "1.0.0"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    df_print: kable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 9, fig.height = 4.5, dpi = 150)
options(stringsAsFactors = FALSE)
```

# Citation

When using the code for **growth categories and birth weight**, **within-animal
growth response**, or **projection to market weight**, please cite:

Pedro Righetti Arnaut, Candido Pomar, Luciano Hauschild, and Aline Remus (2026).
*Increasing the threonine-to-lysine ratio improves the growth performance of
slow-growing pigs within group precision feeding*. Manuscript submitted to the
*Journal of Animal Science*.

**Corresponding author:** [aline.remus@agr.gc.ca](mailto:aline.remus@agr.gc.ca)  
**Additional email:** [alnremus@gmail.com](mailto:alnremus@gmail.com)

## Analytical principle

The analysis follows each pig across growth phases and compares treatment
responses within predefined growth-rate categories. Using each pig's pre-trial
growth as its own baseline helps separate the dietary response from its previous
growth trajectory. Market-weight projections extend the observed growth rates
and use bootstrap resampling to describe their uncertainty; they are not direct
measurements at 120 kg.

Base R only. No packages are needed to run the analyses; `rmarkdown` and `knitr` are
needed only to knit this file.

Two input files, both in the same folder as this document:

* `growth_rates_all_phases.csv` — one row per pig per phase, 59 pigs x 4 phases.
* `feed_cost_by_pig.csv` — one row per pig, feed cost over the 21-d period.

Treatments are conventional phase feeding at 100% of the ideal threonine to lysine ratio of
0.65 (`CPF100`) and group precision feeding at 80, 100, 120 and 140% of that ratio
(`GPF80`, `GPF100`, `GPF120`, `GPF140`). Growth-rate categories (`Gr`) are Fast, Medium, Slow.

# Data

```{r data}
raw <- read.csv("growth_rates_all_phases.csv")
str(raw)
table(raw$Phase)

ORDER <- c("CPF100", "GPF80", "GPF100", "GPF120", "GPF140")
GRP   <- c("Fast", "Medium", "Slow")

# one row per pig, with the quantities used below
wide <- function(ph, cols) {
  s <- raw[raw$Phase == ph, ]
  s <- s[order(s$Pig), cols]
  s
}
b <- wide("Birth-weaning",     c("Pig","Sex","Treatment","Growth","BW_start_kg","BW_end_kg","ADG_kg_d"))
n <- wide("Weaning-selection", c("Pig","BW_end_kg","ADG_kg_d"))
p <- wide("Selection-start",   c("Pig","BW_end_kg","ADG_kg_d"))
t <- wide("Treatment",         c("Pig","BW_end_kg","ADG_kg_d"))

d <- data.frame(
  Pig       = b$Pig,
  Sex       = b$Sex,
  Treatment = factor(b$Treatment, levels = ORDER),
  Gr        = factor(b$Growth,    levels = GRP),
  bw_birth  = b$BW_start_kg,
  bw_wean   = b$BW_end_kg,
  bw_sel    = n$BW_end_kg,
  bw_d1     = p$BW_end_kg,
  bw_end    = t$BW_end_kg,
  adg_suck  = b$ADG_kg_d,
  adg_nurs  = n$ADG_kg_d,
  adg_pre   = p$ADG_kg_d,
  adg_trial = t$ADG_kg_d
)
d$cell  <- interaction(d$Treatment, d$Gr, drop = TRUE)
d$dADG  <- d$adg_trial - d$adg_pre     # within-animal change, kg/d
nrow(d)
```

A convenience function for mean and standard error.

```{r helper}
mse <- function(x) c(mean = mean(x), sem = sd(x) / sqrt(length(x)))
cellmeans <- function(v, f1 = d$Treatment, f2 = d$Gr) tapply(v, list(f1, f2), mean)
```

# Are the growth-rate categories birth-weight categories?

The categories were formed from pre-trial average daily gain. The question is whether they
also separate the pigs by birth weight, that is, whether the slow growers are the
low-birth-weight animals of the population.

```{r birthweight}
round(t(sapply(GRP, function(g) c(
  n          = sum(d$Gr == g),
  birth      = mean(d$bw_birth[d$Gr == g]),
  birth_sem  = sd(d$bw_birth[d$Gr == g]) / sqrt(sum(d$Gr == g)),
  weaning    = mean(d$bw_wean[d$Gr == g]),
  bw_d1      = mean(d$bw_d1[d$Gr == g])
))), 3)

summary(aov(bw_birth ~ Gr, data = d))
summary(aov(bw_wean  ~ Gr, data = d))
summary(aov(bw_d1    ~ Gr, data = d))
```

Correlation of birth weight with each later trait.

```{r corr}
targets <- c(adg_suck = "ADG birth to weaning", adg_nurs = "ADG weaning to selection",
             adg_pre  = "pre-trial ADG",        bw_d1    = "BW at d 1",
             adg_trial= "ADG during the trial")
do.call(rbind, lapply(names(targets), function(k) {
  ct <- cor.test(d$bw_birth, d[[k]])
  data.frame(trait = targets[[k]], r = round(unname(ct$estimate), 3),
             r2 = round(unname(ct$estimate)^2, 3), P = signif(ct$p.value, 3))
}))
```

Cross-classification of birth-weight tertile against the assigned category.

```{r tertile}
cut3 <- quantile(d$bw_birth, c(1/3, 2/3))
d$bw_tertile <- cut(d$bw_birth, c(-Inf, cut3, Inf), labels = c("light","mid","heavy"))
round(cut3, 2)
tab <- table(d$bw_tertile, d$Gr)
tab
chisq.test(tab)
```

# When did the body weight advantage of the GPF140 slow growers appear?

```{r gpf140-slow}
sl <- d[d$Gr == "Slow", ]
round(do.call(rbind, lapply(ORDER, function(tr) {
  s <- sl[sl$Treatment == tr, ]
  c(n = nrow(s), birth = mean(s$bw_birth), weaning = mean(s$bw_wean),
    selection = mean(s$bw_sel), d1 = mean(s$bw_d1))
})), 2) -> tabsl
rownames(tabsl) <- ORDER; tabsl

for (k in c("bw_birth","bw_wean","bw_sel","bw_d1")) {
  f <- summary(aov(sl[[k]] ~ sl$Treatment))[[1]]
  cat(sprintf("%-9s F(%d,%d) = %5.2f   P = %.3f   GPF140 - CPF100 = %+6.2f kg\n",
              k, f$Df[1], f$Df[2], f$`F value`[1], f$`Pr(>F)`[1],
              mean(sl[[k]][sl$Treatment == "GPF140"]) -
              mean(sl[[k]][sl$Treatment == "CPF100"])))
}
```

Average daily gain of the slow growers in each interval, before and during the trial.

```{r slow-adg}
round(do.call(rbind, lapply(ORDER, function(tr) {
  s <- sl[sl$Treatment == tr, ]
  c(suckling = mean(s$adg_suck), nursery = mean(s$adg_nurs),
    pre_trial = mean(s$adg_pre), trial = mean(s$adg_trial))
})), 3) -> tabadg
rownames(tabadg) <- ORDER; tabadg
```

# Within-animal change in average daily gain

Each pig is its own baseline, so a growth trajectory established before the diets were applied
cancels out.

```{r dadg}
round(cellmeans(1000 * d$dADG), 0)          # g/d
round(tapply(1000 * d$dADG, list(d$Treatment, d$Gr),
             function(x) sd(x)/sqrt(length(x))), 0)   # SEM

for (g in GRP) {
  s <- d[d$Gr == g, ]
  f <- summary(aov(dADG ~ Treatment, data = s))[[1]]
  gp <- s[s$Treatment != "CPF100", ]
  ratio <- c(GPF80 = 0.8, GPF100 = 1.0, GPF120 = 1.2, GPF140 = 1.4)[as.character(gp$Treatment)]
  lin <- summary(lm(dADG ~ ratio, data = data.frame(dADG = gp$dADG, ratio = ratio)))
  cat(sprintf("%-7s treatment P = %.3f | linear Thr trend within GPF P = %.3f | GPF140 - CPF100 = %+4.0f g/d\n",
              g, f$`Pr(>F)`[1], coef(lin)[2, 4],
              1000 * (mean(s$dADG[s$Treatment == "GPF140"]) -
                      mean(s$dADG[s$Treatment == "CPF100"]))))
}
```

# Projection to a 120 kg body weight

Nothing was observed at 120 kg. The trial lasted 21 d; what follows is what the measured
average daily gain would imply if it continued unchanged. Two bases are computed.

**Basis 1, cell ADG adjusted for initial body weight.** Average daily gain is adjusted to a
common initial body weight by the pooled within-cell regression of ADG on initial body weight,
and the projection starts from that common weight.

**Basis 2, within-animal change.** For each cell, ADG is the mean pre-trial ADG of the
category plus the mean within-animal change of that cell. Differences among treatments then
come only from the change each animal showed when the diets were applied.

```{r projection-functions}
ancova_slope <- function(df) {
  num <- den <- 0
  for (cl in levels(droplevels(df$cell))) {
    s <- df[df$cell == cl, ]
    num <- num + sum((s$bw_d1 - mean(s$bw_d1)) * (s$adg_trial - mean(s$adg_trial)))
    den <- den + sum((s$bw_d1 - mean(s$bw_d1))^2)
  }
  num / den
}

days_basis1 <- function(df) {                       # days from the END of the period
  B <- mean(df$bw_d1); bta <- ancova_slope(df)
  out <- matrix(NA, length(ORDER), length(GRP), dimnames = list(ORDER, GRP))
  for (tr in ORDER) for (g in GRP) {
    s <- df[df$Treatment == tr & df$Gr == g, ]
    adg <- mean(s$adg_trial) + bta * (B - mean(s$bw_d1))
    out[tr, g] <- (120 - B) / adg - 21
  }
  out
}

days_basis2 <- function(df) {
  out <- matrix(NA, length(ORDER), length(GRP), dimnames = list(ORDER, GRP))
  for (g in GRP) {
    sg   <- df[df$Gr == g, ]
    preg <- mean(sg$adg_pre); bwg <- mean(sg$bw_d1)
    for (tr in ORDER) {
      s   <- sg[sg$Treatment == tr, ]
      adg <- preg + mean(s$dADG)
      out[tr, g] <- (120 - (bwg + 21 * adg)) / adg
    }
  }
  out
}

round(days_basis1(d), 1)
round(days_basis2(d), 1)
```

## Bootstrap

Whole animals are resampled with replacement within each treatment x growth-rate cell. 
```{r bootstrap, cache=TRUE}
set.seed(20260827)
R <- 4000
boot1 <- boot2 <- array(NA, c(R, length(ORDER), length(GRP)),
                        dimnames = list(NULL, ORDER, GRP))
cells <- split(seq_len(nrow(d)), d$cell)
for (i in seq_len(R)) {
  idx <- unlist(lapply(cells, function(ix) sample(ix, length(ix), replace = TRUE)))
  bd  <- d[idx, ]
  boot1[i, , ] <- days_basis1(bd)
  boot2[i, , ] <- days_basis2(bd)
}

contrast <- function(bt) {
  out <- do.call(rbind, lapply(GRP, function(g)
    do.call(rbind, lapply(setdiff(ORDER, "CPF100"), function(tr) {
      v <- bt[, "CPF100", g] - bt[, tr, g]
      data.frame(Gr = g, Treatment = tr, median = round(median(v), 1),
                 lower = round(quantile(v, .025), 1),
                 upper = round(quantile(v, .975), 1), row.names = NULL)
    }))))
  out
}
contrast(boot1)   # basis 1
contrast(boot2)   # basis 2
```

## Figure: projected trajectories

```{r trajectories, fig.width=11, fig.height=3.6}
B   <- mean(d$bw_d1); bta <- ancova_slope(d)
col <- c(Fast = "#d95f02", Medium = "#1f77b4", Slow = "#2ca02c")
pch <- c(Fast = 19, Medium = 15, Slow = 17)
op  <- par(mfrow = c(1, 5), mar = c(4, 1.2, 2.4, 0.6), oma = c(0, 3.4, 0, 0), las = 1)
for (tr in ORDER) {
  plot(NA, xlim = c(-23, 62), ylim = c(B - 2, 128), xlab = "", ylab = "",
       main = tr, yaxt = if (tr == ORDER[1]) "s" else "n")
  abline(h = 120, lty = 2, col = "grey60"); abline(v = 0, col = "grey60")
  for (g in GRP) {
    s   <- d[d$Treatment == tr & d$Gr == g, ]
    adg <- mean(s$adg_trial) + bta * (B - mean(s$bw_d1))
    dd  <- (120 - B) / adg - 21
    lines(c(-21, 0), B + adg * c(0, 21), col = col[g], lwd = 2.4)
    lines(c(0, dd),  B + adg * c(21, dd + 21), col = col[g], lwd = 1.8, lty = 2)
    points(dd, 120, col = col[g], pch = pch[g], cex = 1.2)
  }
  if (tr == ORDER[1]) legend("topleft", GRP, col = col[GRP], pch = pch[GRP],
                             bty = "n", cex = 0.9)
}
mtext("Body weight, kg", side = 2, outer = TRUE, line = 2, las = 0)
mtext("Days from the end of the experimental period", side = 1, outer = TRUE, line = -1.4)
par(op)
```

# Feed cost over the experimental period

Each pig's daily blend priced with 2025 reference ingredient costs. The prices are not current,
so only the comparison among treatments should be read from them.

## Ingredient reference costs — editable upstream inputs

The following values document the ingredient prices used upstream to prepare
`feed_cost_by_pig.csv`. They are shown here to make the assumptions transparent
and easy to revise. **Editing this table alone does not recalculate the results
below**; after changing an ingredient price, the upstream feed-cost calculation
and `feed_cost_by_pig.csv` must also be updated.

Values of `$0.000` for soybean hulls, salt, and choline chloride are placeholders
for missing prices, not evidence that these ingredients have no cost. Until
invoice or reference prices are entered, absolute feed costs may be
underestimated.

| Ingredient | Cost, USD/t | Cost, USD/kg | Source | Source / note | Used in upstream formulas? | Input status |
|:--|--:|--:|:--|:--|:--:|:--|
| Corn | $201.760 | $0.202 | Statistics Canada, 2025 | [Statistics Canada table 32-10-0077-01](https://doi.org/10.25318/3210007701-eng) | Yes | Provided |
| Wheat | $254.060 | $0.254 | Statistics Canada, 2025 | [Statistics Canada table 32-10-0077-01](https://doi.org/10.25318/3210007701-eng) | Yes | Provided |
| Wheat shorts | $108.930 | $0.109 | USDA, 2025 | USDA *National Mill-Feeds and Miscellaneous Feedstuff Report* (as cited in the supplied document) | No | Provided |
| Timothy meal | $335.000 | $0.335 | USDA, 2025 | USDA *National Mill-Feeds and Miscellaneous Feedstuff Report* (as cited in the supplied document) | No | Provided |
| Soybean meal | $408.090 | $0.408 | Statistics Canada, 2025 | [Statistics Canada table 32-10-0077-01](https://doi.org/10.25318/3210007701-eng) | Yes | Provided |
| Limestone | $1,800.000 | $1.800 | Calculated[^calculated-cost] | Calculated by difference of inclusion into experimental diets (supplied Table S1.1 footnote) | Yes | Provided |
| Animal fat | $1,247.600 | $1.248 | Statistics Canada, 2025 | [Statistics Canada table 32-10-0077-01](https://doi.org/10.25318/3210007701-eng) | No | Provided |
| L-Lysine | $1,888.000 | $1.888 | All About Feed, 2025b | [Lysine HCl market price](https://www.allaboutfeed.net/market-prices/lysine-hcl/) | Yes | Provided |
| L-Tryptophan | $6,750.000 | $6.750 | All About Feed, 2025b | All About Feed, 2025b, as supplied; the bibliography in the supplied document points to the L-Lysine HCl page | Yes | Provided |
| DL-Methionine | $3,230.000 | $3.230 | All About Feed, 2025a | [DL-Methionine market price](https://www.allaboutfeed.net/market-prices/dl-methionine/) | Yes | Provided |
| L-Threonine | $2,080.000 | $2.080 | All About Feed, 2025c | [Threonine market price](https://www.allaboutfeed.net/market-prices/threonine/) | Yes | Provided |
| L-Valine | $10,000.000 | $10.000 | All About Feed, 2025b | All About Feed, 2025b, as supplied; the bibliography in the supplied document points to the L-Lysine HCl page | Yes | Provided |
| Vitamin-mineral complex | $10,000.000 | $10.000 | Calculated[^calculated-cost] | Calculated by difference of inclusion into experimental diets (supplied Table S1.1 footnote) | Yes | Provided |
| Monocalcium phosphate | $2,500.000 | $2.500 | Calculated[^calculated-cost] | Calculated by difference of inclusion into experimental diets (supplied Table S1.1 footnote) | Yes | Provided |
| Phytase | $2,000.000 | $2.000 | Calculated[^calculated-cost] | Calculated by difference of inclusion into experimental diets (supplied Table S1.1 footnote) | Yes | Provided |
| Soybean hulls | $0.000 | $0.000 | Not provided | Not available in the supplied reference-cost table | Yes | **Missing — enter invoice/reference price** |
| Salt | $0.000 | $0.000 | Not provided | Not available in the supplied reference-cost table | Yes | **Missing — enter invoice/reference price** |
| Choline chloride | $0.000 | $0.000 | Not provided | Not available in the supplied reference-cost table | Yes | **Missing — enter invoice/reference price** |

[^calculated-cost]: Calculated value supplied in the source document rather
    than a directly observed market quotation.

```{r cost}
fc <- read.csv("feed_cost_by_pig.csv")
fc$Treatment <- factor(fc$Treatment, levels = ORDER)
fc$Growth    <- factor(fc$Growth,    levels = GRP)
fc$per_day   <- fc$Feed_cost_USD_21d / fc$Days

round(tapply(fc$per_day, list(fc$Treatment, fc$Growth), mean), 3)
round(tapply(fc$per_day, fc$Treatment, mean), 3)
round(100 * tapply(fc$per_day, fc$Treatment, mean) /
      mean(fc$per_day[fc$Treatment == "CPF100"]), 1)     # relative to CPF100 = 100
round(tapply(fc$Thr_cost_USD_21d, fc$Treatment, mean), 3)
round(100 * tapply(fc$Thr_cost_USD_21d, fc$Treatment, mean) /
      tapply(fc$Feed_cost_USD_21d, fc$Treatment, mean), 1)   # % of feed cost
```

# Session information

```{r session}
sessionInfo()
```
